---
title: "R Basics"
output: github_document
---

# 1. BASIC FUNCTIONS & DATA SORTING ####
## 1a. The Foundations of R  ====
### 1.1a. Getting Started ----

```{r Headers}

# Headers can be used to organize a script for data wrangling, manipulations, and analyses

# 1. LEVEL 1 HEADER ####
## 1a. Level 2 Header ====
### 1.1a. level 3 header. ----

```

```{r Directories}

getwd() # Shows current working directory file path

setwd("/User/Richard/Desktop") # Sets new working directory based on file path input

getwd() # Double check to make sure your working directory is set correctly

```

```{r Packages}

install.packages("tidyverse") # installs tidyverse package
install.packages("tidyverse", "dslabs") # installs  multiple packages at once

installed.packages() # Checks to see what packages are installed

library(tidyverse) # loads a library
library(tidyverse, dslabs) # loads multiple libraries at once

# Once a package is installed on your machine, you will only need to load it for when you open R next time

```

```{r Loading_Data}

library(readr) # load the readr library
library(R.matlab) # load the R.matlab library


Data <- read.csv("filepath/filename.csv",
                 header=T) # load CSV files

Data <- read.table("filepath/filename.txt", 
                   header=T) # load .txt files

Data <- read.table("filepath/filename.txt", 
                   header=T, sep="\t") 
# \t is a separator for tabs. R converts missing values into "NA"

Data <- readMat("filepath/filename.mat") # load .mat files

# Header = TRUE (capitalized) if you have headers in the dataset
# TRUE can be abbreviated as T
# Header = FALSE in the read.table command tells R that the first row
# of the dataset is actual data

# Loading data from the internet
# Type the URL of the dataset instead of the location of the file
read.table("http://people.math.umass.edu/~lavine/Book/book.data/pinecones.txt", header=T)
pinecones<-read.table("http://people.math.umass.edu/~lavine/Book/book.data/pinecones.txt", header=T) 

print(pinecones)
head(pinecones)

```

```{r Cleaning_Up}

rm(murders) # removing one object from the Environment (workspace)

rm (murders, abb) # removing more than one object

rm(list = ls()) # clear environment from packets and dataset (Ctrl+Shift+F10)

dev.off()  # clear plots

cat("\014")  # clear console

```

### 1.2a. Objects and Functions in R ----

```{r Simple_Math}

# R can serve as a simple calculator 

# Type any number below
9

# Add two numbers
6 + 3

# Divide two numbers
9/3

# Multiply two numbers
4.5*3

# Square (or other) a number
3^2

# Conduct multiple operations on one line
6; 6 + 3; 9/3; 4.5*3; 3^2

```

```{r Objects}

# Anything in R is treated as an "object"
# Objects can be variables, functions, a statistical model, or even a whole data set

# Data types can be characters (strings), numerical or integers (non-decimal numbers),
# or logical operators (True or False)

# Object names should be meaningful, not include spaces (use "_" instead). 
# and should not contradict other object or function names

print ("I am an Object?!") # character (string) object
4.5 + 4.5 # numerical object
1:10 #  vector of numerical values from 1 to 10
6 + 3 # integer object 

# This assigns a variable/value to an object
x = 3
y = 6
print(x) # shows variable
print(y) # shows variable

z = x > y  # tests a logical argument and assigns the outcome to an object
print(z) # shows variable

class(z) # prints the object class of an object

ls() # Shows names of all objects saved in work space

```

```{r Functions}

# Functions are pre-built code that performs specific operations based on inpit arguments 
# Parentheses are needed to evaluate any function, and often require at least 1 argument

# Assigns a value to a function
log(8)

# Assigns an already defined object to a function
a <- 1
log(a)

# Functions follow order of inside to out layering of parentheses
log(exp(1)) # First, does exp of 1, then takes the log of that

# Call help files to read over function manual
help("log")

# Show argument requirements for a function
args(log)

# R assumes arguments are entered by the order input is entered unless specified otherwise

```

### 1.3a. Data Types ----

```{r DataFrames}

# Data Frames are most common way of storing datasets 

# Useful for combining multiple types of data into one object

library(dslabs) # load librarys
data("murders") # loads murders dataset

class(murders) # checks to see what type of data the input is

str(murders) # shows structure of the object
names(murders) # provides column names of the object
head(murders) # shows first 6 lines of dataset

# Rows often represent cases/observations
# Columns often represent variables

# Accessor Operator obtains information from a given column
murders$population # Shows all values for this column

# Head returns the first or last parts of a vector, matrix, table, data frame or function
# Tail displays the last few rows
head(murders, n=7)
tail(murders, n=10)

```

```{r Creating_Tables}

# Make a table for the number of species living in a region
Wildlife <- c(rep("Wolvering", 250), # first input indicates name/value, second is reps
            rep("Grey Wolve", 670),
            rep("Deer", 7337),
            rep("Elk",  3770),
            rep("Bald Eagle", 855),
            rep("Woodpecker", 2369),
            rep("Golde Eagle", 157),
            rep("California Condor", 97))
print(Wildlife)

# Create a frequency table for these data showing total counts
freqs <- table(Wildlife)
print(freqs)

```

```{r Creating_DataDrames}

# Creating a data frame with stringAsFactors = FALSE
grades <- data.frame(names = c("John", "Juan", "Jean", "Yao"), # one columns
                     exam_1 = c(95, 80, 90, 85), # another column
                     exam_2 = c(90, 85, 85, 90), # another column
                     stringsAsFactors = FALSE) # column names with strings are NOT made into factors
grades 

# Creating a dataframe from already existing vectors

names<- c("James", "Marcos", "Elena", "Andrea")
ages<- c(52, 15, 42, 35)
weight<- c(160, 155, 165, 158)
Stats<-data.frame(Name = names, Age = ages, Weigh = weight)
print(Stats)

is.data.frame(Stats) # checks to see if object is a dataframe


# Other ways to create data frames
m<-rnorm(10, 2, 3)
m
n<-rnorm(10, 5, 2)
n
o<-data.frame(m, n)
o

k <- data.frame(x = rnorm(20, 5, 2.3), y = runif(20, 3, 4.5))
k

# Creating a matrix as an object containing only one type of variable
s<- matrix(1:9, nrow=3) 
s
v<-matrix(1:16, nrow=4)
v
g<- matrix(1:25, nrow=5)
g
t<-matrix(0, nrow =7, ncol=7)
t
is.matrix(t)

```

```{r Vectors}

# A vector is an object containing several inputs which can vary in length

# Define a new vector
a<-1+1 # creates a vector of one unit (the nunber 2)
print(a) # shows variable

b<-21:30  # creates a vector of numbers from 21 to 30
print(b) # shows variable

c<-c(1, 5, 9, 13, 12, 15)  # a vector containing several numbers concatenated 
print(c) # shows variable

d<-rep(c(5,10),3)   # vector repeating the numbers 5 and 10 for 3 times 
print(d)# shows variable

e <- f <- g <- 1992  # creates a cascade of assignments for a vector

h <- c("US", "UK", "Germany") # vector of characters

# Matrix operations based on the previous vector values
a*b
a/b
a^b
a+b
a-b
aa<-a+b
aa
2*a
a/3

# Indexing vectors
c[1] # examines first element of the vector
c[2] # examines second element of the vector
c[3] # examines third element of the vector

c[3:6] # examines the third to sixth elements of the vector
c[c(3,6)] # examines the third and sixth elements of the vector

c[-3] # examines all but the third element of the vector
c[-c(3, 6)] # examines all but the third and sixth elements of the vector

murders [,] # accessing all rows and columns
murders[1, ] # accessing first row and all columns
murders[1:3,] # accessing first three rows and all columns
murders[ c(1,3), ] # accessing first and third row and all columns
murders[ , 3] # accessing column 3 and all rows
murders[1, 2:3] # accessing first row and 2nd-3rd column
murders[1:2, 2:3] # accessing 1st-2nd row, 2nd-3rd column
murders[,c(1,3)] # accessing 1st and 3rd column with all rows

murders$population[murders$population>10000] # accessing a subset of data
unique(murders$state) # accessing nmber of unique values in a variable

# Using existing data to create vectors
population <- murders$population # creates numerical vector
length(population) # shows number of observations for that vector

# Characters can also be stored as vectors
class(murders$state) # shows character vector in dataframe, often indicated with "s

# Logical vectors are either TRUE or FALSE
z <- 22 == 22.5
z
class(z) # Puts out true or false output

# Factors are categorical in data
class(murders$region)

levels(murders$region) # Shows levels of factor

# Checking the vector type
is.vector(a) # examine if the object is a vector
is.numeric(a) # examine if the object is a numerical vector

# Levels are stored as integers in R, but this is mostly for efficiency 

sequences <- seq(1,10) # vector from 1 to 10
sequences <- seq(1,10, 3) # vector from 1 to 10 for every 3rd number

# Applying arithmetic functions to vectors
aa <- c(1, 2, 4, 7, 10)
aa
sum(aa)
length(aa)
m <- sum(aa)/length(aa) # m is the arithmetic mean
m

# Creating the arithmetic mean by making a function
Mean<- function(x){ m <- sum(x)/length(x); return(m) }
Mean(aa)

# Another example
bb <- 1:10
bb
Mean(bb)
mean_bb <- sum(bb)/length(bb) # basic mean calculation

# Evaluating the structure of vectors and datasets
ls() # view all objects created in the workspace

str(murders) # shows structure of dataset
dim(murders) # number of dimensions
names(murders) # names of the columns of dataset
levels(murders$state) # levels in a factor (categorical) variable

```

## 1b. Vector Manipulation  ====
### 1.1b. Sorting by Vectors ----

```{r Vector_Coercion}

# Vectors must be all of the same type

x<- c(1, "canada",3) # R converted the 1 and 3 into characters to match the "Canada" string

# Specific coercions can be forced
x <- 1:5 # vector of 1 to 5
y <- as.character(x) # converts vector of x into character strings

# NAs can be introduced from coercion in some cases like below
x<- c("1", "b", "3")
as.numeric(x)

# NAs are encountered often, so be ready to handle that

```

```{r Vector_Sorting}

library(dslabs) # loads library
data(murders) # loads data

# Data can be arranged using sort, order, or rank

# Sort will take a vector and arrange it from the minimum to maximum value
# (e.g., x <- c(10, 5, 15), sort will result in: 5, 10, 15)

# Order will create an index of the sorted data
# (e.g., x <- c(10, 5, 15), order will result in 2, 1, 3)

# Rank will provide a vector of ranks for original data
# (e.g., x <- c(10, 5, 15), Rank will result in 2, 1, 3)

# Sorting data
sort(murders$total) # sorts data in ascending order

# Order function will give index of vector
x <- c(31, 4, 15, 92, 65) # create simple vector
x
sort(x) # sorts vector in ascending order
index <- order(x) # returns index that will put x in order
x[index] # rearranging by this index puts elements in order
order(x) # shows index that puts x in order

# Order state names by total murders
murders$state[1:10]
murders$abb[1:10]

index <- order(murders$total) # Define the index as the order of total murders
print(index)
murders$abb[index] # Change the order of the abbreviations column based on the total murders index (showing states lowest to most murders)  

max(murders$total) # compute the highest number from this column
i_max <- which.max(murders$total) # index the highest number of murders from this column
murders$state[i_max] # list the name with the highest number of murders based on the index location 

i_min <- which.min(murders$total)
murders$state[i_min]

# Rank ordering
x <- c(31, 4, 15, 92, 65)
x
rank(x)    # returns ranks (smallest to largest)

```

```{r Vector_Operations}

# Can sort column based on values from other columns
murders$state[which.max(murders$population)] # Shows which state has the biggest population

# Can compute new data from existing columns
murder_rate <- murders$total / murders$population * 100000 # Compute murder rate 
print(murder_rate)

# Sort a column based on decreasing order of another column ordering the states by murder rate
murders$state[order(murder_rate, decreasing=TRUE)]

```

### 1.2b. Indexing Data  ----

```{r Logical_Operators}

murder_rate <- murders$total / murders$population * 100000 # defines murder rates in a new column
index <- murder_rate <= 0.71 # logical vector indicating if the murder rate is less than or equal to 0.71
index

murders$state[index] # shows which states have a lower murder rate than 0.71
sum(index) # shows how many states have less than 0.71 murder rates

# Logical Vectors can represent Conditions 
west <- murders$region == "West" # logical vector indicating which states are in the west
west

safe <- murder_rate <= 1 # logical vector indicating which states are safe (i.e., less than or equal to 1 murder rate)
safe

index <- safe & west # index that considers when both logical operators are true
index
murders$state[index] # shows which states in the west are safe

```

```{r Indexing_Functions}

# Which provides entries of a logical vector that are true

# Match searches for entries in a vector and returns their index

# %in% will examine if each element of a first vector is within a second vector

# Indexing with Which
x <- c(FALSE, TRUE, FALSE, TRUE, TRUE, FALSE) # create logical vector
which(x) # shows index of which values are true

index <- which(murders$state == "Massachusetts") # index the state of Massachusetts
index
murder_rate[index] # look at murder rate for that state based on the index

# Indexing with Match
index <- match(c("New York", "Florida", "Texas"), murders$state) # index the states that match the input
index
murders$state[index]
murder_rate[index]

# Indexing with the %in% operator
x <- c("a", "b", "c", "d", "e")
y <- c("a", "d", "f")
y %in% x # provides logical outcomes as to whether the letters in y are also in x

c("Boston", "Dakota", "Washington") %in% murders$state # shows if these names are located in the states column

```

```{r Indexing_DataFrames}

# You can also index specific aspects (rows/columns) from a dataframe

library(dslabs) # load librarys
data("murders") # loads murders dataset

Viewmurders # to display the data set in a new tab in the data editor window
head(PlantGrowth) # Render the headers and first 6 rows
murders[1,1]  # Render the value located in cell (Row,Column). Row=Column=1
murders[1,2]  # Render the value located at 1st row and 2nd column
murders[2,1]  # Render the value located at 2nd row and 1st column
murders[5,]   # Render header and all values from the 5th row
murders[,2]   # Render all values from the 2nd column

murders$state # The values of state are displayed

# Displaying variables' values without using the command "data$variable" by defining the variables as objects
state<-murders[,1]  # Render all values from the 1st column
state                   

abb<-murders[,2]   # Render all values from the 2nd column
abb

```

### 1.3b. Data Wrangling  ----

```{r Basic_Wrangling}

install.packages("dplyr") # install dplyr package
library(dplyr) # load dplyr library
library(dslabs) # load other libraries

data("murders") # load data

# Mutate function will add new columns or changing existing columns

# Filter function will subset data by specific rows

# Select function will subset data by specific columns

# Pipe Operator will send one function to another 

# Mutate to create new columns
murders <- mutate(murders, rate = total / population * 100000) # creates a new column
head(murders)

# Filter to create subset rows
filter(murders, rate <= 0.71) # grab only rows with a rate less than or equal to 0.71

# Select only specific columns of data
new_table <- select(murders, state, region, rate) # only grab data from state, region, and rate columns
head(new_table)

# Pipe Operator to combine this all
murders %>% # specify the data frame
  select(state, region, rate) %>% # grab the specified columns from the main data frame
  filter(rate <= 0.71) # subset to only look at rows with rates lower than or equal to 0.71

```

## 1c. Simple Plots  ====
### 1.1c. Basic R Plots  ----

```{r Base_R_Plots}

library(dplyr, dslabs) # load libraries
data("murders") # load data

# Basic scatterplot
x <- murders$population /10^6
y <- murders$total
plot(x, y)

# Basic histogram
murders <- mutate(murders, rate = total / population * 100000)
hist(murders$rate)

# Basic boxplot
boxplot(rate~region, data = murders)

```

## 1d. Summarizing Data ====
### 1.1d. Basic Summarization Functions  ----

```{r Summarization}

# Basic summaries can include mean, median, standard deviation, and standard error
# Summarize function computes these summary statistics for data

library(tidyverse, dplyr, dslabs) # load libraries
data(murders) # load data
murders <- mutate(murders, rate = total / population * 10^5) # adjusted murder rate

# Summarize Function

# Compute the min, median, and max murder rates for states in the west region
Stats <- murders %>% # send murders data down the pipe operator
  filter(region == "West") %>% # only grab region with "west" 
  summarize(minimum = min(rate),  # grab min from all rows
            median = median(rate), # grab median from all rows
            maximum = max(rate)) # grab max from all rows
print(Stats) # thisis stored as a table now

Stats$median # accessing the component median with the accessor $
Stats$minimum # accessing the component min with the accessor $
Stats$maximum # accessing the component max with the accessor $

s$median
s$maximum

# Compute the average murder rate for all states 
mean(murders$rate)

# Compute the average murder rate for all states adjusted by population size
us_murder_rate <- murders %>%  # send murders data down the pipe operator
  summarize(rate = sum(total) / sum(population) * 10^5)  # compute population adjusted murder rate and get the mean from that (i.e., sum(total)/sum(population)) 
print(us_murder_rate)

```

```{r Summarization_Quantiles}

# Quantiles function can also extract the min, max, and median values

library(tidyverse, dplyr, dslabs) # load libraries
data(murders) # load data
murders <- mutate(murders, rate = total / population * 10^5) # adjusted murder rate

# Quantiles takes a vector and returns the specified values (0 = min, 0.5 = median, 1 = max)
quantile(murders$total, c(0, 0.5, 1))

# Quantiles with the Summarize Function 
Stats <- murders %>%  # send murders data down the pipe operator
  filter(region == "West") %>%  # only grab region with "west" 
  summarize(range = quantile(rate, c(0, 0.5, 1))) # quantile function

# Returning Quantiles as a dataframe
my_quantile <- function(x){
  r <-  quantile(x, c(0, 0.5, 1))
  data.frame(minimum = r[1], median = r[2], maximum = r[3]) # specifies that min is in column 1, median in column 2, and max in column 3
}
murders %>%  # send murders data down the pipe operator
  filter(region == "West") %>% # only grab region with "west" 
  summarize(my_quantile(rate)) # quantile function
print(my_quantile)

```

```{r Pull_Function}

# Pull function will return a given vector as a numerical value

library(tidyverse, dplyr, dslabs) # load libraries
data(murders) # load data
murders <- mutate(murders, rate = total / population * 10^5) # adjusted murder rate

# Compute the average rate adjusted by population size
us_murder_rate <- murders %>%  # send murders data down the pipe operator
  summarize(rate = sum(total) / sum(population) * 10^5) # computes mean murder rate across all states
print(us_murder_rate) # average murder rate is stored as a dataframe

# Pull function to return average murder rate as a numeric value
us_murder_rate %>% pull(rate)

# Pull function with Summarize function to save the number directly
us_murder_rate <- murders %>% # send murders data down the pipe operator
  summarize(rate = sum(total) / sum(population) * 10^5) %>% # computes mean murder rate across all states
  pull(rate) # Apply pull function to this new value
print(us_murder_rate) # average murder rate is now stored as a number

```

```{r Dot_Placeholder}

# Dot Placeholders (.) serves as a placeholder for data being passed through the pipe 

library(tidyverse, dplyr, dslabs) # load libraries
data(murders) # load data
murders <- mutate(murders, rate = total / population * 10^5) # adjusted murder rate

# Compute the average rate adjusted by population size
us_murder_rate <- murders %>%  # send murders data down the pipe operator
  summarize(rate = sum(total) / sum(population) * 10^5) # computes mean murder rate across all states
print(us_murder_rate) # average murder rate is stored as a dataframe

# Use the Dot Placeholder 
us_murder_rate <- murders %>% # send murders data down the pipe operator
  summarize(rate = sum(total) / sum(population) * 10^5) %>% # computes mean murder rate across all states
  .$rate # Apply dot placeholder
print(us_murder_rate)

class(us_murder_rate) # the dot place holder served as a pull function here

```

```{r Summarization_by_Groups}

# Group_by function allows for data to be separated into groups

library(tidyverse, dplyr, dslabs) # load libraries
data(murders) # load data
murders <- mutate(murders, rate = total / population * 10^5) # adjusted murder rate

# Group by region
murders %>% group_by(region)

# Summarize after grouping
Stats <- murders %>% # send murders data down the pipe operator
  group_by(region) %>% # applies group by function
  summarize(median = median(rate)) # grab median from each group
print(Stats)

```

```{r Sorting_Tables}

# Arrange function can change the order of an entire dataframe or table

library(tidyverse, dplyr, dslabs) # load libraries
data(murders) # load data
murders <- mutate(murders, rate = total / population * 10^5) # adjusted murder rate

# Arrange the order of states by population size
murders %>% arrange(population) %>% head()

# Arrange the order of states by murder rate - the default is ascending order
murders %>% arrange(rate) %>% head()

# Arrange the order of states by murder rate in descending order
murders %>% arrange(desc(rate)) %>% head()

# Arrange the order of states by region and then by murder rate within region
murders %>% arrange(region, rate) %>% head()

# Return the top 10 states by murder rate
murders %>% top_n(10, rate)

# Return the top 10 states ranked by murder rate, sorted by murder rate
murders %>% arrange(desc(rate)) %>% top_n(10)

```

## 1e. Working with Data Table Function ====
### 1.1e. Data Table  ----

```{r Data_Table}

# Data Table package is a powerful package that can serve as an alternative to dplyr

install.packages("data.table") # install packages
library(data.table) # load libraries
library(tidyverse, dplyr, dslabs) # load libraries
data(murders) # load data

# Creating Data Table from existing dataframe
murders <- setDT(murders)
class(murders)

# Selecting Columns using Data Table
murders[, c("state", "region")] |> head() # option 1
murders[, .(state, region)] |> head() # option 2

# In option 2 the .() indicates that this is a column name 

# Adding or Changing Columns using Data Table
murders[, rate := total / population * 100000] # compute adjusted murder rate 
head(murders)

# the := allows for updating columns by reference

# Create Multiple Columns using Data Table
murders[, ":="(rate = total / population * 100000, rank = rank(population))] # 

# y is referring to x and := changes by reference
x <- data.table(a = 1)
y <- x

x[,a := 2] # If x changes, then y will also be changed
y

y[,a := 1]
x

# Use copy to make an actual copy
x <- data.table(a = 1)
y <- copy(x)
x[,a := 2]
y

```

```{r Subsetting_in_Data_Table}

# Subsetting data can also be done in Data Table similar to dplyr

install.packages("data.table") # install packages
library(data.table) # load libraries
library(tidyverse, dplyr, dslabs) # load libraries
data(murders) # load data

# Prepare Data for Examples
murders <- setDT(murders) # Assign as data table
murders[, rate := total / population * 100000] # adjusted murder rate

# Subsetting in Data Table
murders[rate <= 0.7] # only show rows with murder rate below or equal to 0.7

# Combining Filter and Select in Data Table
murders[rate <= 0.7, .(state, rate)] # only show rows with murder rate below or equal to 0.7, and then grab the columns state and rate 

```

### 1.2e. Summarizing Using Data Table  ----

```{r Summarization_Data_Table}

install.packages("data.table") # install packages
library(data.table) # load libraries
library(tidyverse, dplyr, dslabs) # load libraries
data(heights) # load data
heights <- setDT(heights) # set as Data Table

# Summarizing using Data Table
stats <- heights[, .(average = mean(height), 
                     standard_deviation = sd(height))]
print(stats)

# Subsetting and then summarizing using Data Table
stats <- heights[sex == "Female", 
                 .(average = mean(height),
                   standard_deviation = sd(height))]
print(stats)

# Create quantile function for min, median, and max
median_min_max <- function(x){
  qs <- quantile(x, c(0.5, 0, 1))
  data.frame(median = qs[1], minimum = qs[2], maximum = qs[3])
}

# Apply quantile function using Data Table
stats <- heights[, .(median_min_max(height))]
print(stats)

# Apply group_by function using Data Table
stats <- heights[, .(average = mean(height), 
            standard_deviation = sd(height)), 
            by = sex]
print(stats)

```

```{r Sorting_Data_Table}

install.packages("data.table") # install packages
library(data.table) # load libraries
library(tidyverse, dplyr, dslabs) # load libraries
data(murders) # load data
murders <- setDT(murders) # set as Data Table
murders[, rate := total / population * 100000] # compute murder rate based on population

# Change the order of data by population using Data Table
murders[order(population)] |> head()

# Change the order of data by population using Data Table in descending order
murders[order(population, decreasing = TRUE)] 

# Change the order of data by region then by murder rate using Data Table
murders[order(region, rate)]

```

# 2. PROGRAMMING BASICS ####

## 2a. Conditionals & Functions  ====
### 2.1a. Conditional Expressions ----

```{r Basic_Conditional_Expressions}

# Conditional expressions assist in sorting through data

# ifelse() works like an if-else statement on each elemeent of a vector

# any() works with a vector of logicals and returns true if any elements are true

# all() works with a vector of logicals and returns true if all elements are true

# An if-else statement example
a <- 0 # assigns a as 0

if(a!=0){ # if a does not = 0
  print(1/a) # print the recirpricol of a
} else{
  print("No reciprocal for 0.") # or print the string if it is 0
}

# ifelse() general form
# if(boolean condition){
#   expressions
# } else{
#   alternative expressions
# }

# Examine which states have murder rate less than 0.5

library(dslabs) # loads library
data(murders) # loads data

murder_rate <- murders$total / murders$population*100000 # calculate murder rate in vector
ind <- which.min(murder_rate) #identify minimal murder rate index

# Conditional statement looking if the lowest ind is < 0.5
if(murder_rate[ind] < 0.5){
  print(murders$state[ind]) 
} else{
  print("No state has murder rate that low")
}

# Conditional statement looking if the lowest ind is < 0.25
if(murder_rate[ind] < 0.25){
  print(murders$state[ind]) 
} else{
  print("No state has a murder rate that low.")
}

# ifelse function is similar to if-else conditional
a <- 0
ifelse(a > 0, 1/a, NA) # ifelse combines this all in one line

# vector example
a <- c(0,1,2,-4,5)
result <- ifelse(a > 0, 1/a, NA) # if item in vector is greater than 0, 1/it, or list as NA

# Using ifelse to identify and replace NAs with 0s
data(na_example) # load data
sum(is.na(na_example)) # Compute sum of all NAs in data

no_nas <- ifelse(is.na(na_example), 0, na_example) # if NA, replace with 0, or the same value
sum(is.na(no_nas)) # Compute sum of all NAs

# the any() and all() functions evaluate logical vectors
z <- c(TRUE, TRUE, FALSE) # logical vector
any(z) # checks if any of the logical vector contain true
all(z) # cehcks if all of the logical vector contain true

```

### 2.1b. Functions ----

```{r Basic_Functions}

# Functions can be the bread and butter post-workout meal you've been craving if applied right

# Define a function
avg <- function(x){ # create function name and inputs
  s <- sum(x) # define inner function aspects
  n <- length(x) # define inner function aspects
  s/n # define inner function aspects to produce output
}

# All variables defined within a function are not saved to the workspace 
s <- 3 
avg(1:10) # 5.5 is not saved on workspace
s

# Function general form
# my_function <- function(VARIABLE_NAME){
#   perform operations on VARIABLE_NAME and calculate VALUE
#   VALUE is the final line in a function
# }

# Functions with multiple arguments

avg <- function(x, arithmetic = TRUE){ # Only apply function if arithmetic vector
  n <- length(x) # find length of vector
  ifelse(arithmetic, sum(x)/n, prod(x)^(1/n)) # if not arithmetic, result in geometric mean
}

```

### 2.1c. For Loops ----

```{r For_Loops}

# For-loops define a range of variables to perform given operations on (e.g., each subject in a file, each row in a dataset, etc.)


# Function computing sum of integers 1 through n
compute_s_n <- function(n){
  x <- 1:n
  sum(x)
}

# Function applied to 1:55 
compute_s_n(55)

# Basic For-Loop (i.e., for integers 1 to 5, print the integer)
for(i in 1:5){
  print(i)
}

# For-Loop applies to function
m <- 30 # create variable
s_n <- vector(length=m) # creates an empty vector up to the size of m

for(n in 1:m){ # for 1 through the size of m
  s_n[n] <- compute_s_n(n) # fill in the current vector index with the sum of all integers up to that point
}

# Plot these results
n <- 1:m # create a vector for 1 to size of m
plot(n, s_n) # plot the vector by the summation values
lines(n, n*(n+1)/2)

# Create a table of values from formula to compare if function in for-loop produces same results
head(data.frame(s_n = s_n, formula = n*(n+1)/2))

# Add lines from formula to predicted plot
plot(n, s_n) + lines(n, n*(n+1)/2)

```






