---
title: "Basic Data Checks"
output: github_document
---

# 1. DATA PREPARATIONS ####
## 1a. Set Up  ====
### 1.1a. Install Packages & Load Libraries ----

```{r Packages}

# Install the following packages
install.packages("psych")     
install.packages("tidyverse") 
install.packages("readr")  

```

```{r Libraries}

# Load the following libraries
library(psych)
library(tidyverse)
library(readr)

```

### 1.2a. Set Up Directory & Load Data ----

```{r Directory}

# Directory
getwd()
setwd("/Users/wardr/Desktop")
getwd()

```

```{r Load_Data}

# Load data
library(dslabs) 
data("murders") # loads murders dataset

```

## 1b. Dealing with Duplicate Cases & Anomolies ====
### 1.1b. Identify & Remove Duplicate Cases ----

```{r Duplicate_Cases}

# Look at data
view(murders)

# Identify duplicates in data
duplicated(murders) # Identifying cases that are duplicates, returns "TRUE" for cases repeating
murders[duplicated(murders), ]  # Only displaying duplicates

```

```{r Unique_Cases}

# Look at data
view(murders)

# Identify unique cases in data
unique(murders) # Lists unique cases

```

```{r Dealing_with_Duplicates}

# Look at data
view(murders)

# Create New Dataset Without Duplicates or With Only Unique Cases
NewMurders1 <- unique(murders)
NewMurders2 <- distinct(murders)

# Look over new dataset (note no differences between the ways we did this)
view(NewMurders1)
view(NewMurders2)

```

### 1.2b. Identify & Remove Anomolies ----

```{r Detect_Anomolies}

# Summary statistics of dataset to identify outliers
summary(murders)

# Sort categorical variables
murders_StateSorted <- murders[order(murders$state), ] # Sorts state by ascending order
murders_StateAbbSorted <- murders[order(!murders$abb), ] # Sorts state abbreviation by ascending order
murders_Region <- murders[order(murders$region), ] # Sorts region by ascending order

# Look over data
view(murders_StateSorted)
view(murders_StateAbbSorted)
view(murders_Region)

```

```{r Correct_Anomolies}

# DC is listed as being in the "South", let's change this to "Northeast"
DC_indx <- which(murders$abb == "DC") # index DC
murders[DC_indx, "region"] = "Northeast"

```

## 1c. Dealing with Missing Values ====
### 1.1c. Identify Missing Values ----

```{r Identify_Missing}

# Identify indices of missing values for each column
which(is.na(murders$region))

# Calculate missing cases for a column
AllMiss <- is.na(murders$region) # Creating a vector for the missing values with logical returns 
sum(AllMiss) # Total counts of missing values

# Calculate Proportion of Missing Data from All Columns
PropMiss <- function(x){sum(is.na(x))/length(x)*100} # function to calculate % of missing cases
apply(murders, 2, PropMiss) # Calculate % of missing cases in all columns

# Calculate Proportion of Non-Missing Data from All Columns
PropPresent <- function(x){sum(!is.na(x))/length(x)*100} # function to calculate % of non-missing cases
apply(murders, 2, PropPresent) # Calculate % of non-missing cases in all columns

```

### 1.2c. Remove Cases with Missing Values ----

```{r Removing_Missing}

# Create new datset without missing cases
No_NA_Murders <- na.omit(murders) # new data creating omitting NAs

# Or you can apply the na.rm = "T" function to a dataframe
mean(murders$total, na.rm = "T")

#   *******  Creating a new data set without missing values   ******* 

sum(is.na(training2))           # Total No. of missing values in Training2
newtraining<-na.omit(training2) # Creating the data set without missing values
sum(is.na(newtraining))         # Checking whether newtraining has missing values

```

### 1.3c. Working with Missing Values ----

```{r Correlations_with_Missing}

# Run correlations using only complete pairs
cor(murders$population, murders$total, use="pairwise.complete.obs", method="pearson")    

```

```{r Replace_Missing}

# Replace missing values with a new number or 0
Variable <- murders$region # create vector for column
sum(is.na(Variable)) # calculate all missing cases
Variable[is.na(Variable)] = 0 # replace with 0 for one variable
murders[is.na(murders)] = 0 # replace with 0 for all variables with missing cases

# Can also use "ifelse" logic to replace missing values with mean of a variable
Variable <- ifelse(is.na(Variable), 74.73, Variable)

```

## 1d. Outlier Identification & Removal ====
### 1.1d. Identifying Outliers ----

```{r BoxPlot_2_Identify_Outliers}

# Basic Box Plot
boxplot(murders$population) # creates boxplot
boxplot.stats(murders$population) # provides statistics showing lower whisker, 25th, 50th, 75th, and upper whisker ($stats), count ($n), confidence intervals ($conf) and outliers ($out) from boxplot 

```

## 1e. Normalize Variables ====
### 1.1e. Basic Normalization Techniques ----

```{r Scale_Function}

# Z-score data using scale function
z_score <- scale(murders$population)

```

```{r Basic_Normalizations}

# Z-score data using scale function
z_score <- scale(murders$population)

# Min/Max Normalization 
minPop <- min(murders$population) # minimum value
maxPop <- max(murders$population) # maximum value
rangePop <- maxPop - minPop # calculate the range of variable
normPop <- (murders$population - minPop)/rangePop # normalize variable based on min and max
summary(normPop)

```
